### 2019-09-24 manuscript.tex
edits to the Broad tectonic history overview section

Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have been a central constituent of both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography. There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models.

@@

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces as the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016a}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+Nain provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). This model posits that this period terminal collision is not only associated with the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the Midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny which is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Whitmeyer2007a}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces. This plutonism is likely due to crustal melting within a back-arc region of ca. 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}. Younger magmatic activity ca. 1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}.

@@

Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca. 1.25–1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11-1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces. This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen. The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.

There was significantly less crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}) and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca. 1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated. \citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep[e.g.]{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting. Along the western margin of Laurentia, small-scale rifting occurred ca. 780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Rooney2017a}. However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran (closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary; \citealp{Bond1984a, Levy1991a}). The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin. If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up. One possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence. The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin. Available geochronological constraints and thermal subsidence modeling indicate ca. 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland at the time.

Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a,Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this  history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Ca. 620-580 Ma rifting initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane. Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz–Arequipa–Rio Apa (MARA) block MARA block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Famatinian orogeny.

The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis. As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian one in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian period) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval). These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision. In the case of the Grenville interval this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated to with the assembly of the supercontinent Pangea.

>>>>

Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography. There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models. This chapter seeks to provide a concise review of these records.

@@

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne+Nain provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave province and the Rae province ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and Nain provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the Nain province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks was not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016a}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+Nain provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny. Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic \citep{Whitmeyer2007a}. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occurred ca. 1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces extending from the southwest United States up to the Central Gneiss Belt of Ontario to the northeast of Georgian Bay \citep{Slagstad2009a}. This plutonism is likely due to crustal melting within a back-arc region of ca. 1.50 to 1.43 Ga accretionary orogenesis \citep{Bickford2015a}. Younger magmatic activity ca. 1.37 Ga  of the Southern Granite–Rhyolite Province suggests a similar tectonic setting of accretionary orogenesis at that time \citep{Bickford2015a}. While an active margin interpretation with magmatism in back‐arc setting has gained traction within the literature with evidence, the tectonic setting is considered enigmatic given earlier interpretations of an anorogenic setting (see references in \citealp{Slagstad2009a}).

@@

Accretionary orogenesis continued along the (south)east margin of Laurentia with the arc-continent collision of the ca. 1.25-1.22 Ga Elzevirian orogeny \citep{McLelland2013a}. The subsequent ca. 1.19 to 1.16 Ga Shawinigan orogeny is interpreted to be due to the accretion of a terrane comprised of amalgamated arc volcanics and associated metasediments and is followed by a period of tectonic quiescence on the eastern margin of Laurentia until the collision orogenesis of the Grenvillian orogeny \citep{McLelland2010a}. In the latest Mesoproterozoic (ca. 1.11-1.08 Ga), a major intracontinental rift co-located with a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior province and Paleoproterozoic provinces. This Midcontinent Rift lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism through the orogen \citep{McLelland2010a}. The orogeny is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a}.

There was significantly less crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}) and the Mesoproterozoic tectonic history is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with ca. 1.47 to 1.40 intracontinental rift -- the tectonic setting of which is debated. \citet{Hoffman1989a} proposed that it may be a remanent back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep[e.g.]{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting. Along the western margin of Laurentia, small-scale rifting occurred ca. 780 to 720 Ma leading to deposition in basins that is recorded from the Death Valley region of SW Laurentia up to the Mackenzie Mountains of NW Laurentia \citep{Rooney2017a}. However, this extensional basin development is relatively minor and predates the more significant rifting that lead to passive margin thermal subsidence that did not occur until the Ediacaran (closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary; \citealp{Bond1984a, Levy1991a}). The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin. If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports active tectonism along the western margin of Laurentia, but suggests that more dramatic lithospheric thinning occurred later than the timing of rifting typically implemented in models of Rodinia break-up. One possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence. The margin that did experience large-scale rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin. Available geochronological constraints and thermal subsidence modeling indicate ca. 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland ca. 820 Ma.

Extensive rifting that was followed by thermal subsidence occurred along the southeast to east Laurentia margin leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a,Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Ca. 620-580 Ma rifting initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phases of this rifting are associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. Cuyania is widely interpreted be a rifted fragment of SE Laurentia that separated associated with this early Cambrian rifting and subsequently became part of Gondwana when it collided with other terranes in the vicinity of the Rio de Plata craton during the Ordovician Famatinian orogeny \citep{Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting of a major craton as the record that lingers on the craton is similar. One interpretation is that there was successful break-up along the eastern margin during the ca. 580 to 550 Ma interval of rifting prior to the ca. 539 Oklahoma aulacogen rifting that liberated the Cuyania microcontinent. The Maz–Arequipa–Rio Apa (MARA) block MARA block with which Cuyania collided \citep{Martin2019a} is likely a product of such rifting. Orogenesis between the MARA block and the Rio de Plata and Kalahari in the ca. 530 Ma Pampean orogeny \citep{Casquet2018a} predated the collision of Cuyania during the ca. 460 Famatinian orogeny.

The eastern margin of Laurentia then went through the cycle of Appalachian orogenesis. As is visualized in Figure \ref{fig:tectonic_history}, there are parallels between the Grenville orogenic interval and the Appalachian orogenic interval in that there was a period of arc-continent collision (Shawinigan orogeny in the Grenville interval; Taconic orogeny in the Appalachian interval) followed by microcontinent accretion (Llano in the Grenville interval; Acadian in the Appalachian interval) that culminated in large-scale continent-continent collision (Grenvillian orogeny in the Grenville interval; Alleghanian in the Appalachian interval). These similarities are the consequence of an active margin facing an ocean basin that was progressively consumed until its consumption resulted in continent-continent collision. In the case of the Grenville interval, this terminal collision is interpreted to be associated with the assembly of the supercontinent Rodinia and in the Appalachian interval it is interpreted to be associated to with the assembly of the supercontinent Pangea.
